---
title: "Advanced Data Visualization with R at JH"
description: |
  The sequel in the JH dataviz specialization
author:
  - name: David Harper
    url: https://dh-data.org/
date: 2022-08-27
output:
  distill::distill_article:
    self_contained: false
---

```{r include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)

library(tidyverse)

# echo =    FALSE: evaluates the code, hides the code, and shows output in the knit
# include = FALSE: evaluates the code, hides the code, but DOES NOT SHOW the output in the knit

```

This three-week course follows Data Visualization in R with ggplot2 (I previously [shared my final submission](https://dh-data.org/posts/2022-06-06-my-jh-dataviz-submission/) to that prerequisite) and is also taught by the talented Collin Paschall. Despite being dubbed advanced, I perceived the difficulty to be low to moderate, if you are proficient with R; e.g., the peer-reviewed assignments are *easier* than those in the prerequisite course. The strength of the course is that it gave me practice in plots that otherwise I'd never make. The weakness is a relative lack of theory. I actually expected this course to go deeper into the grammatical theory, but don't expect that here. For theory, I tend to rely on [ggplot2: elegant graphics for data analysis](https://ggplot2-book.org/), and obviously the indispensibele [reference](https://ggplot2.tidyverse.org/index.html). The course refers to [Data Visualization with R](https://rkabacoff.github.io/datavis/) and [R Graphics Cookbook](https://r-graphics.org/). Below is a sample (< 50%) of the plots.

### Extensions of scatterplots 

Although it's easy to add a line to a scatterplot, there is an entire library of possibilities under the surface. To the geom_point() we only need to add geom_smooth() where the default is [locally weighted polynomial; aka, LOESS](https://stat.ethz.ch/R-manual/R-devel/library/stats/html/loess.html). When adding a line to a scatterplot, a key choice is between the more familiar linear model (e.g., "method = lm") and the default. As illustrated by my fourth plot below, the lm method can plot a polynomial. I collected four plots with the [patchwork package](https://patchwork.data-imaginist.com/index.html) and it is indeed "ridiculously simple" to use!

```{r include=FALSE}

set.seed(19)
var1<-runif(50,0,10)
var2<-var1+rnorm(50,5,2)
var3<- var1^2*(-rnorm(50,1,.2))
dat1<-tibble(var1,var2)

```

```{r include=TRUE}

library(patchwork) 

p1 <- dat1 %>% ggplot(aes(x = var1, y = var2)) +
    geom_point() +
    geom_smooth() + 
    labs(x = "Variable1", y = "Variable2")

p2 <- dat1 %>% ggplot(aes(x = var1, y = var2)) +
    geom_point() +
    geom_smooth(span = 0.3) + 
    labs(x = "Variable1", y = "Variable2")

p3 <- dat1 %>% ggplot(aes(x = var1, y = var3)) +
    geom_point() +
    geom_smooth(method = lm, se = FALSE) + 
    labs(x = "Variable1", y = "Variable2")

p4 <- dat1 %>% ggplot(aes(x = var1, y = var3)) +
    geom_point() +
    geom_smooth(method = lm, formula = y ~ poly(x, 2), se = FALSE) + 
    labs(x = "Variable1", y = "Variable2")

lm_4 <- lm(var3 ~ poly(var1, 2), data = dat1)
lm_4$coefficients

(p1 + p2)/ (p3 + p4) # patchwork layout magic with simple arithmetic

```

<br>
A scatterplot matrix is acheived with the GGally which [extends ggplot2](https://ggobi.github.io/ggally/index.html). Notice that ggpairs() generates a matrix where I have defined the diagonal and lower triangle with ggplot wrapper functions.

```{r include=FALSE}

library(GGally)

set.seed(10)
var1<-runif(50,0,10)
var2<-var1+rnorm(50,5,2)
var3<- var1*(-rnorm(50,1,.2))

dat2<-tibble(var1,var2,var3)

```

```{r include=TRUE}
my_density <- function(data, mapping, ...){
  ggplot(data = data, mapping = mapping) + 
    geom_density(fill = "#0433ff", ...)
}

my_scatter <- function(data, mapping, ...){
  ggplot(data = data, mapping = mapping) + 
    geom_point(size = 2,
               color = "#ffa500")
}

ggpairs(dat2,
        diag = list(continuous = my_density),
        lower = list(continuous = my_scatter))
```

<br>
I was reminded that a typical barplot is not an efficient use of ink. This is a lollipop plot; it is similar to a [Cleveland dot plot](https://r-graphics.org/recipe-bar-graph-dot-plot). It uses geom_segment() for the stem, and it uses geom_point() for the circle.

```{r problem_5_data,include=FALSE}

#####DO NOT MODIFY THIS CHUNK

set.seed(13)
var1<-rnorm(10,25,5)
names<-letters[1:10]

dat5<-tibble(names,var1)
```

```{r problem_5,echo=FALSE}

####Make sure you load any necessary libraries

dat5 %>% ggplot(aes(x = reorder(names, var1), y = var1)) + 
    geom_point(size = 3, color = "#006400") + 
    geom_segment(x = names, xend = names, y = 0, yend = var1, color = "#006400", size = 1) + 
    ylim(0,35) +
    theme(axis.text.x = element_text(angle = 90)) +
    labs(x = NULL, y = "Variable 1")

```

<br>
The correlation matrix is popular and useful for EDA. This courses uses [ggcorrplot](https://rpkgs.datanovia.com/ggcorrplot/index.html). To illustrate the package, I decided to retrieve cryptocurrency prices for the top dozen (in volume) coins. Notice how I used pivot_wider() to render the dataframe into a wide format that is friendly to the cor() function.

```{r}
library(ggcorrplot)
library(lubridate)

# install.packages("crypto2")
# library(crypto2)
# coins_top12 <- crypto_list(only_active = TRUE) %>% arrange(rank) %>% head(12)
# coin_hist <- crypto_history(coins_top12, start_date = "20220101", end_date ="20220131")
# saveRDS(coin_hist, file = "coin_hist.rds")

coin_hist <- readRDS(file = "coin_hist.rds")
coin_hist2 <- coin_hist %>% 
    mutate(day_n = mday(timestamp)) %>% 
    select(day_n, slug, close) %>% 
    pivot_wider(id_cols = day_n, names_from = slug, values_from = close) %>% 
    select(-day_n)
coin_hist2_r <- cor(coin_hist2)
coin_hist2_r %>% ggcorrplot(type = "lower", 
                            show.diag = TRUE,
                            hc.order = TRUE,
                            lab = TRUE, lab_size = 2,
                            colors = c("coral2", "azure3", "chartreuse2"))

```

### Geospatial visualization 

My favorite was the second week's introduction to geospatial visualization. For reference, [Hadley's chapter](https://ggplot2-book.org/maps.html) is excellent. This is a choropleth with geom_polygon(). Basic maps can be generated with the maps package and geom_polygon().

```{r problem_2_data,include=FALSE}

library(maps)
set.seed(15)
Measurement<-rnorm(32,50,1)
my_world_map <- map_data("world")
countries <- unique(my_world_map$region)
set.seed(987)
some_data_values <- data.frame(
  "region"=countries,
  "Score"=runif(252,0,100))

my_world_map_2 <- left_join(my_world_map,some_data_values,by="region")

```

Here is a bubble map with geom_polygon() + geom_point()

```{r solution_p2, include=TRUE, message=FALSE}

# library(maps)
# my_world_map wrangled in dis-included chunk

my_cities <-maps::world.cities
latin_cities <- my_cities %>% filter(
        country.etc == "Costa Rica" |
        country.etc == "Panama" |  
        country.etc == "Nicaragua") %>% filter(
        pop > 40000) %>% add_column(Measurement)

my_world_map %>% filter(region == "Costa Rica" | region == "Panama" | region == "Nicaragua") %>%
    ggplot(aes(x = long, y = lat, group = group)) +
    geom_polygon(color="black",fill="white") + 
    geom_point(data = latin_cities,aes(x=long, y=lat, group=NULL, color = Measurement), size = 5) +
    scale_color_distiller(palette=7)

```

<br>
A better way is to use geom_sf() from the awesome [Simple Features for R](https://r-spatial.github.io/sf/index.html).

```{r problem_3_data,include=FALSE}

library(sf)
library(rnaturalearth)
library(rnaturalearthdata)
library(rgeos)

s_america<-ne_countries(scale="medium",continent='south america',returnclass="sf")

```

```{r problem3,echo=FALSE}
# dis-included packages include {rnaturalearth}
# s_america<-ne_countries(scale="medium",continent='south america',returnclass="sf")

ggplot() + geom_sf(data = s_america, aes(fill = pop_est)) +
    scale_fill_distiller(palette = 10)

```


### gganimate and plotly

Plotly is powerful in ways I don't understand, but is also can simply convert a ggplot2 object. This was the third week. I hope that's an interesting tour of what you'd learn in this course.

```{r problem_3_2_data,include=FALSE}

Response<-c("Energize","Amazing","Great")
set.seed(9819)
Energize<-tibble(Company=rep("Energize",100),Output=rnorm(100,50,20))
set.seed(9819)
Amazing<-tibble(Company=rep("Amazing",100),Output=rnorm(100,50,10))
set.seed(9819)
Great<-tibble(Company=rep("Great",100),Output=rnorm(100,40,5))

my_dat<-bind_rows(Energize,Amazing,Great)

```

```{r problem_2_solution,include=TRUE,message=FALSE,warning=FALSE}

library(plotly)

my_boxplot <- my_dat %>% ggplot(aes(x = Company, y = Output, fill = Company)) + 
    geom_boxplot() +
    theme(legend.position = "right")

ggplotly(my_boxplot)

```

I hope that's helpful and interesting. I almost forgot: the [course page is here](https://www.coursera.org/learn/jhu-advanced-data-visualization-r)



