---
title: "BT Question Set P1-T2-20-18: Multivariate regressions"
description: |
  Fama-French three-factor model; House prices; and Medical costs
author:
  - name: David Harper
    url: https://www.bionicturtle.com/
date: 2022-01-08
output:
  distill::distill_article:
    self_contained: false
---

### Multiple regression

In each exercise, the values aspire to be realistic but, becasue I wanted to test for specific items, the datasets are simulated. 

### Question 1: Fama-french

20.18.1. Sally is a portfolio manager at an investment management firm. She wants to test her primary equity portfolio's reaction to the factors in the Fama-French three-factor model. She collected excess returns (i.e., net of the riskfree rate) over the last eight years, so that the sample size, n = 96 months. The response (aka, explained, dependent) variable is the portfolio's excess return. The three explanatory variables are the market factor (MKT), the size factor (SMB), and the value factor (HML). The size factor captures the excess return of small capitalization stocks (SMB = "small minus big") and the value factor captures the excess returns of value stocks (HML = "high book-to-market minus low book-to-market"). Sally's regression results are displayed below.

_(regression table here)_

Which of the following descriptions of her portfolio is the most accurate?

a. Her small capitalization, value-oriented low-beta portfolio has not generated alpha
b. Her large capitalization, growth-oriented high-beta portfolio has not generated alpha
c. Her large capitalization, growth-oriented low-beta portfolio has generated significantly positive alpha
d. Her small capitalization, value-oriented high-beta portfolio has generated significantly positive alpha

```{r warning=FALSE, message=FALSE}

library(tidyverse)
library(broom)
library(gt)

intercept <- .03
intercept_sig <- .01

x1_mu <- .04
x1_sig <- .01
x1_beta <- 0.4

x2_mu <- .03
x2_sig <- .01
x2_beta <- -0.6

x3_mu <- .03
x3_sig <- .01
x3_beta <- -0.3

noise_mu <- 0
noise_sig <- 0 # low value gets low p-value b/c low noise

size <- 96
set.seed(18)

results <- tibble(
  x0 = rnorm(size, intercept, intercept_sig),
  x1 = rnorm(size, x1_mu, x1_sig),
  x2 = rnorm(size, x2_mu, x2_sig),
  x3 = rnorm(size, x3_mu, x3_sig),
  x1_b = rep(x1_beta, size),
  x2_b = rep(x2_beta, size),
  x3_b = rep(x3_beta, size),
  noise = rnorm(size, 0, noise_sig)
)


results1 <- results %>% mutate(
  y = x0 +x1_b * x1 + x2_b * x2 + x3_b * x3 + noise
)

model <- lm(y ~ x1 + x2 + x3, data = results1)
summary(model)

model_tidy <- tidy(model)
model_tidy[2,1] <- "MKT"
model_tidy[3,1] <- "SMB"
model_tidy[4,1] <- "HML"

gt_table_model <- gt(model_tidy)

gt_table_model <- 
  gt_table_model %>% 
  tab_options(
    table.font.size = 14
  ) %>% 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body()
  ) %>% 
  tab_header(
    title = "Portfolio excess returns regressed against MKT + SMB + HML",
    subtitle ="i.e., Fama-French three-factor model"
  #) %>% tab_source_note(
  #   source_note = md("the source is ... FRED")
  ) %>% cols_label(
    term = "Coefficient",
    estimate = "Estimate",
    std.error = "Std Error",
    statistic = "t-stat",
    p.value = "p value"
  ) %>% fmt_number(
    columns = vars(estimate, std.error, statistic, p.value),
    decimals = 3
  ) %>% fmt_scientific(
    columns = vars(statistic, p.value),
  ) %>% tab_options(
    heading.title.font.size = 14,
    heading.subtitle.font.size = 12
  )

gt_table_model

```

### Question 2: House Prices

20.18.2. Derek regressed house prices (as the response or dependent variable) against three explanatory variables: square footage (SQFEET), number of rooms in the house (ROOMS), and age of the house (AGE). The dependent variable, PRICE, is expressed in thousands of dollars ($000); e.g., the average PRICE is $386.051 because the average house price in the sample of 96 houses is $386,051. The units of SQFEET are unadjusted units; e.g., the average SQFEET in the sample is 1,203 ft^2. The variable ROOMS is equal to the sum of the number of bedrooms and bathrooms; because much of the sample is 2- and 3-bedroom houses with 2 baths, the average of ROOM is 4.55. Finally, AGE is given in years where the average AGE in the sample is 14.77 years. Derek's regression results are displayed below.

_(regression table here)_

Each of the following statements is true about these regression results EXCEPT which is false?

a. Older houses have lower prices on average
b. The 98.0% confidence interval (CI) for the AGE coefficient is (5.7, 10.4)
c. The 90.0% confidence interval (CI) for the ROOMS coefficient is (8.1, 10.9)
d. An additional (+) 100 square feet (ft^2) is associated with an expected increase of ~ $29,100 in the price of the house

```{r warning=FALSE, message=FALSE}

library(tidyverse)
library(broom)
library(gt)

intercept <- 40
intercept_sig <- .01

x1_mu <- 1200
x1_sig <- 30
x1_beta <- 0.35

x2_mu <- 4.5
x2_sig <- 2
x2_beta <- 10.0

x3_mu <- 15
x3_sig <- 4
x3_beta <- -8.0

noise_mu <- 0
noise_sig <- 20 # low value gets low p-value b/c low noise

size <- 96
set.seed(43)

results <- tibble(
  x0 = rnorm(size, intercept, intercept_sig),
  x1 = rnorm(size, x1_mu, x1_sig),
  x2 = rnorm(size, x2_mu, x2_sig),
  x3 = rnorm(size, x3_mu, x3_sig),
  x1_b = rep(x1_beta, size),
  x2_b = rep(x2_beta, size),
  x3_b = rep(x3_beta, size),
  noise = rnorm(size, 0, noise_sig)
)


results1 <- results %>% mutate(
  y = x0 +x1_b * x1 + x2_b * x2 + x3_b * x3 + noise
)

model <- lm(y ~ x1 + x2 + x3, data = results1)
summary(model)

model_tidy <- tidy(model)
model_tidy[2,1] <- "SQFEET"
model_tidy[3,1] <- "ROOMS"
model_tidy[4,1] <- "AGE"

gt_table_model <- gt(model_tidy)

gt_table_model <- 
  gt_table_model %>% 
  tab_options(
    table.font.size = 14
  ) %>% 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body()
  ) %>% 
  tab_header(
    title = "House Price regressed against ft^2 (SQFEET) + ROOMS(#) + AGE(years)",
    subtitle = md("House Price in Thousands **($000)** of dollars")
  #) %>% tab_source_note(
  #   source_note = md("the source is ... FRED")
  ) %>% cols_label(
    term = "Coefficient",
    estimate = "Estimate",
    std.error = "Std Error",
    statistic = "t-stat",
    p.value = "p value"
  ) %>% fmt_number(
    columns = vars(estimate, std.error, statistic, p.value),
    decimals = 3
  ) %>% fmt_scientific(
    columns = vars(p.value),
  ) %>% tab_options(
    heading.title.font.size = 14,
    heading.subtitle.font.size = 12
  )

gt_table_model

mean(results1$y) # price
mean(results1$x0) # intercept
mean(results1$x1) # sqfeet
mean(results1$x2) # rooms
mean(results1$x3) # age


```

### Question 3: Insurance

20.18.3. Mary works for an insurance company and she has regressed medical costs (aka, the response or dependent variable) for a sample of patients against four independent variables: AGE, BMI, SMOKER, and CHARITY. The sample's average age is 38.51 years. Body mass index (BMI) is mass divided by height squared and the sample's average BMI is 22.16. SMOKER is a dummy variable where zero indicates a non-smoker and 1 indicates a smoker; the sample's average SMOKER value is 0.163 which indicates that 16.3% of the sample are smokers. CHARITY is the dollar amount of charitable spending in the last year; the sample average is $490.70 donated to charity in the last year. Mary's regression results are displayed below.

_(regression table here)_

Each of the following statements is true about these regression results EXCEPT which is false?

a. The sample size is 43 patients
b. Mary can reject a null hypothesis that all explanatory variables (jointly) have zero coefficients
c. Mary can infer that patient medical cost is positively associated with each of AGE, BMI, and, on average, is greater for a smoker
d. Mary should suspect problematic multicollinearity because the intercept is suspiciously negative and the adjusted R-squared is too near to the unadjusted R-squared

```{r warning=FALSE, message=FALSE}

library(tidyverse)
library(broom)
library(gt)

intercept <- 150
intercept_sig <- 40

# age
x1_mu <- 38
x1_sig <- 7
x1_beta <- 50

# bmi
x2_mu <- 22
x2_sig <- 4
x2_beta <- 100

# smoker
x3_mu <- 15
x3_sig <- 0.5
x3_beta <- 535

# spend
x4_mu <- 500
x4_sig <- 250
x4_beta <- -0.4


noise_mu <- 0
noise_sig <- 300 # low value gets low p-value b/c low noise

size <- 43
set.seed(12)

results <- tibble(
  x0 = rnorm(size, intercept, intercept_sig),
  x1 = round(rnorm(size, x1_mu, x1_sig)),
  x2 = rnorm(size, x2_mu, x2_sig),
  x3 = round(runif(size)-.35), # smoker = 1, non = 0
  x4 = rnorm(size, x4_mu, x4_sig),
  x1_b = rep(x1_beta, size),
  x2_b = rep(x2_beta, size),
  x3_b = rep(x3_beta, size),
  x4_b = rep(x4_beta, size),
  noise = rnorm(size, 0, noise_sig)
)


results1 <- results %>% mutate(
  y = x0 +x1_b * x1 + x2_b * x2 + x3_b * x3 + x4_b * x4 + noise
)

model <- lm(y ~ x1 + x2 + x3 + x4, data = results1)
summary(model)

model_tidy <- tidy(model)
model_tidy[2,1] <- "AGE"
model_tidy[3,1] <- "BMI"
model_tidy[4,1] <- "SMOKER"
model_tidy[5,1] <- "CHARITY"

gt_table_model <- gt(model_tidy)

gt_table_model <- 
  gt_table_model %>% 
  tab_options(
    table.font.size = 14
  ) %>% 
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_body()
  ) %>% 
  tab_header(
    title = "Medical COST regressed against AGE + BMI + SMOKER(1/0) + CHARITY($)",
    subtitle = md("Simulated dataset")
  ) %>% tab_source_note(
    source_note = md("Residual standard error: 295.2 on 38 degrees of freedom")
  ) %>% tab_source_note(
    source_note = md("Multiple R-squared:  0.8343,	Adjusted R-squared:  0.8168")
  ) %>% tab_source_note(
    source_note = md("F-statistic: 47.82 on 4 and 38 DF,  p-value: 2.486e-14")
  ) %>% cols_label(
    term = "Coefficient",
    estimate = "Estimate",
    std.error = "Std Error",
    statistic = "t-stat",
    p.value = "p value"
  ) %>% fmt_number(
    columns = vars(estimate, std.error, statistic, p.value),
    decimals = 2
  ) %>% fmt_scientific(
    columns = vars(p.value),
  ) %>% tab_options(
    heading.title.font.size = 14,
    heading.subtitle.font.size = 12
  )


gt_table_model

mean(results1$y) # cost
mean(results1$x0) # intercept
mean(results1$x1) # age
mean(results1$x2) # bmi
mean(results1$x3) # smoker
mean(results1$x4) # charity

```

