---
title: "My JH dataviz submission"
description: |
  A short description of the post.
author:
  - name: David Harper
    url: https://www.bionicturtle.com/
date: 2022-06-06
output:
  distill::distill_article:
    self_contained: false
---

### Instructions 

Your submission should include the following:

* Using ggplot, create a scatter plot, a distribution figure (box plot, histogram, or density plot), bar plot, and line plot.
* Use colors and facetting in at least one of your figures.
* Add a text annotation to at least one of the your figures.
* Use a ggtheme for one of your figures.
* Make sure your figures have titles and appropriately labeled axes.


My Notes:

1. Scatterplot (Exercise #3), distribution (Exercise #4), Bar plot (Exercise #1), and Line plot (Exercise #2)
2. Exercise #3 includes facet and colors; i.e., I used rgb() to achieve desired color effect
3. The first chart contains an annotation ("Bar height is sample size") suppported by a colored rectangle
4. ggtheme is used in Exercise #1, specifically theme_few()
5. All plots have titles and x/y labels

Following the directions on the Coursera assignment page, you will make four original visualizations. Note that the data for the CCES and CEL data are imported in code in the R Markdown file.

### Put your name here: David R Harper

## Exercise 1

I'm proud of this because it's a bar chart but the (rotated) bar widths represent sample size. The cces data is grouped by Marital Status (i.e., marital). I inverted the pew_region with mutate(pew_religion = 5 - pew_religimp) so that 4 = Very Important. I was curious which groups are the most religious. The highest average is Separated but per the "skinny" bar it's a small sample. It took some work to get it right; e.g., I had to reorder the factors.

Put your figure here:

```{r exercise1}

library(tidyverse)
library(ggthemes)
library(forcats)

cces <- read_rds("cces_dh.rds")
cel <- read_rds("cel_dh.rds")

cces <- cces %>% mutate(pew_religion = 5 - pew_religimp)

cces_sum <- cces %>% 
    group_by(marstat) %>% 
    summarize(
        belief = mean(pew_religion),
        believers = n()) %>% 
    arrange(belief)

# Inspired by https://r-graph-gallery.com/81-barplot-with-variable-width.html 

cces_sum$marital <- recode(cces_sum$marstat, 
       `1` = "Married", 
       `2` = "Separated", 
       `3` = "Divorced",
       `4` = "Widowed", 
       `5` = "Never Married", 
       `6` = "Civil Partner") 

cces_sum$marital <- fct_reorder(cces_sum$marital, cces_sum$belief, .desc = TRUE)

# Calculate the future positions on the x axis of each bar (left border, central position, right border)
cces_sum$marstat <- as.factor(cces_sum$marstat)
cces_sum$right <- cumsum(cces_sum$believers) + 20*c(0:(nrow(cces_sum)-1))
cces_sum$left <- cces_sum$right - cces_sum$believers
 
# Plot
cces_sum %>% ggplot(aes(ymin = 0)) + 
    geom_rect(aes(xmin = left, xmax = right, ymax = belief, colour = marital, fill = marital)) +
    ggtitle("Religiosity grouped by Marital Status") + 
    xlab("Number of respondents") + 
    ylab("Importance of Religion (1 = Not at all Important, ..., 4 = Very important)") +
    coord_flip() +
    theme_few() +
    theme(legend.title = element_blank()) + 
    annotate(geom = "rect", xmin = 640, ymin = 0.9, xmax = 710, ymax = 2.1, fill = "#bfffd0") +
    annotate(geom = "text", x = 680, y = 1.5, label = "Bar height is sample size")

```

## Exercise 2

This is a relatively simple geom_line() with ggtheme theme_tufte. I did color Latinos purple to attempt to _respectfully_ match the purple in the Flag of the Hispanic People. As expected, the number of Congresspersons among these two groups grew over the period.

Put your figure here:

```{r exercise2}

yearly_tbl <- cel %>% group_by(year) %>% 
    summarise(
        "African Americans" = sum(afam),
        Latinos = sum(latino),
        total = n()
    )

# yearly_tbl %>% ggplot(aes(x = year)) + 
#    geom_line(aes(y = no_afam)) + 
#    geom_line(aes(y = no_latino))

yearly_tbl %>% 
    pivot_longer(-c(year, total)) %>% 
    ggplot(aes(x = year, y = value, color = name)) + 
    geom_line(size = 2) +
    scale_color_manual(values = c("dark grey", "purple")) + 
    theme_tufte() + 
    theme(
        text = element_text(size = 14, face = "bold"),
        legend.title = element_blank()
    ) +
    ggtitle("Number of Congresspersons over the years") +
    xlab("Year") +
    ylab("Number within Ethnic Group")

  
```

## Exercise 3

This plot explores whether more tenured congresspersons are more effective. However, I filtered on correlations above 0.40. The colors are custom selected.

Put your figure here:

```{r exercise3}

cel <- cel %>% group_by(st_name) %>% 
    mutate(
        seniority_les = cor(seniority, les)
    ) 

cel$state_full <- state.name[match(cel$st_name, state.abb)]

cel %>% filter(seniority_les > .40) %>% 
    ggplot(aes(x = seniority, y = les)) + 
    geom_jitter(color = rgb(.5, .5, .2), alpha = 0.3) + 
    geom_smooth(method = "lm", se = FALSE, color = rgb(1, .6, 0), size = 1) + 
    labs(
        title = "Seniority versus effectiveness", 
        subtitle = "(Only states where correlation exceeds 0.40)",
        x = "Seniority (Terms in Congress)", 
        y = "Legislative effective score (LES)") + 
    theme(
        plot.title = element_text(size = 14),
        axis.title = element_text(face = "bold", size = 12),
        strip.text.x = element_text(face = "bold", size = 11),
        strip.background = element_rect(fill = rgb(.95, .95, .95)),
        panel.background = element_blank()
    ) + 
    facet_wrap(~state_full) 

```

## Exercise 4

I was curious if the distribution (percent of vote won) appears to be different by party. So these are overlapping histogram. My colors are carefully chosen (via hex) to match each party's colors and achieve a reddish PURPLE where they overlap. 

Put your figure here:

```{r exercise4}

# cel$female <- as.factor(cel$female)

# cel$female <- as.factor(recode(cel$female, `0` = "Male", `1` = "Female"))
# cel$majority <- as.factor(recode(cel$majority, `0` = "Minority", `1` = "Majority"))

cel$dem <- as.factor(recode(cel$dem, `0` = "Republican", `1` = "Democrat"))

cel %>% rename(Party = dem) %>% 
    ggplot(aes(x=votepct, fill = Party)) + 
    geom_histogram(binwidth = 5, color="#e9ecef", alpha=0.5, position = 'identity') +
    scale_fill_manual(values=c("#00AEF3", "#E81B23")) +
    labs(
        title = "Distribution of percent won by party",
        x = "Percent of Vote Won", 
        y = "Number of Congresspersons") + 
    theme(
        panel.background = element_blank(),
    )

```


